### 2023-02-20 tex/gen-sem.tex
[ build ] get thesis into building state
[ writing ] lots of background, before feedback on later chapters
[ writing ] new draft of mechanisation chapter

\def\SimpleSem{../agda/latex/SimpleSem.tex}

@@

With the aim of abstracting away from explicit contexts, bringing us closer to
natural deduction, we can use some new notation to rephrase these requirements.
We will work in \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} rather than \AgdaPrimitive{Set}.
One of the basic connectives in this setting is the \emph{pointwise} arrow
\AgdaFunction{\_$\Rightarrow$\_}, which acts in
\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
like the non-dependent arrow does in \AgdaPrimitive{Set}.
Another basic component is the \AgdaFunction{$\forall[\_]$} notation, which
embeds \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} into \AgdaPrimitive{Set} by using an implicit $\Pi$-type
to quantify over \emph{all} contexts.
Finally, at this stage, I introduce a modality \AgdaFunction{$\bigcirc$}
encapsulating the pattern of considering arbitrary \emph{extensions} of a
context.
To facilitate working in this point-free setting, I give infix versions of
the families \AgdaBound{$\V$} and \AgdaBound{$\C$} (respectively
\AgdaFunction{\_$\V\vdash$\_} and \AgdaFunction{\_$\C\vdash$\_}).
The principal use of these aliases is to fill the right argument with a type
(occuring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
a context given through the point-free machinery.

@@

>>>>

\def\SimpleSem{../agda/processed-latex/SimpleSem.tex}

@@

With the aim of abstracting away from explicit contexts, bringing us closer to
natural deduction, we can use some new notation to rephrase these requirements.
We will work in \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} rather than \AgdaPrimitive{Set}.
One of the basic connectives in this setting is the \emph{pointwise} arrow
\AgdaFunction{\_$\dotto$\_}, which acts in
\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
like the non-dependent arrow does in \AgdaPrimitive{Set}.
Another basic component is the \AgdaFunction{$\forall[\_]$} notation, which
embeds \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} into \AgdaPrimitive{Set} by using an implicit $\Pi$-type
to quantify over \emph{all} contexts.
Finally, at this stage, I introduce a modality \AgdaFunction{$\bigcirc$}
encapsulating the pattern of considering arbitrary \emph{extensions} of a
context.
To facilitate working in this point-free setting, I give infix versions of
the families \AgdaBound{$\V$} and \AgdaBound{$\C$} (respectively
\AgdaFunction{\_$\V\vDash$\_} and \AgdaFunction{\_$\C\vDash$\_}).
The principal use of these aliases is to fill the right argument with a type
(occuring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
a context given through the point-free machinery.

@@

With \AgdaRecord{Semantics} fixed, I also give the corresponding implementation
of \AgdaFunction{sem}.
Like \AgdaFunction{trav} from \cref{sec:syntactic-kits}, we need a
\AgdaFunction{bindEnv} function, but I have updated this to deal with renamings
and the \AgdaFunction{$\Box$}-operator, and also to fit better with concepts
important for generic \emph{syntax} (as in \cref{sec:gen-syn}).
\AgdaFunction{bindEnv} now takes the old environment \AgdaBound{$\rho$}%
\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}\AgdaFunction{[}\AgdaSpace{}%
\AgdaBound{$\V$}\AgdaSpace{}\AgdaFunction{]}\AgdaSpace{}\AgdaBound{$\Gamma$}%
\AgdaSpace{}\AgdaFunction{$\Rightarrow^e$}\AgdaSpace{}\AgdaBound{$\Delta$}, a
renaming \AgdaBound{r}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
\AgdaFunction{Ren}\AgdaSpace{}\AgdaBound{$\Gamma^{+}$}\AgdaSpace{}%
\AgdaBound{$\Gamma$}, and an environment \AgdaBound{$\sigma$} in the renamed
context \AgdaBound{$\Gamma^{+}$} targeting the context extension
\AgdaBound{$\Delta$r}.
In the \AgdaInductiveConstructor{lam} case of \AgdaFunction{sem}, the
\AgdaField{$\llbracket$lam$\rrbracket$} field gives us exactly the renaming to
pass to \AgdaFunction{bindEnv}, while the extension environment
\AgdaBound{$\sigma$} is made as a singleton environment containing just the
\AgdaBound{A}-value \AgdaBound{v}.

\ExecuteMetaData[\SimpleSem]{sem}
\ExecuteMetaData[\SimpleSem]{bindEnv}

### 2023-02-20 tex/gen-syn.tex
[ writing ] new draft of mechanisation chapter

In our generic version of \AgdaRecord{Semantics}, we keep the
\AgdaField{ren\textasciicaron$\V$} and \AgdaField{$\llbracket$var$\rrbracket$}
fields as before, and replace \AgdaField{$\llbracket$app$\rrbracket$} and
\AgdaField{$\llbracket$lam$\rrbracket$} by a
\AgdaField{$\llbracket$con$\rrbracket$} field as follows.
\[
  \AgdaField{$\llbracket$con$\rrbracket$}~\AgdaBound{r}~\AgdaSymbol{:}~%
  \forallb{%
    \Box\plr{{} \env\V \Delta_1 \dotto {} \sdtstile{}\C A_1} \dottimes
    \cdots \dottimes
    \Box\plr{{} \env\V \Delta_n \dotto {} \sdtstile{}\C A_n} \dotto
    {} \sdtstile{}\C A}
\]
I use ${} \sdtstile{}\C A$ for the Agda notation
\AgdaFunction{\_$\C\vDash$}\AgdaSpace{}\AgdaBound{A}, while ${} \env\V \Delta$
stands for the type family of environments
\AgdaSymbol{$\lambda$}\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}%
\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaFunction{Env}\AgdaSpace{}\AgdaBound{$\V$}%
\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaBound{$\Delta$}.
Environments appear in this definition simply as a way to write a product of
$\V$-values --- one value for each element of $\Delta$.
We could make a special case of premises which do not bind any variables, as did
\citet{AACMM21}, eliding the $\Box$ and empty environment, but I choose not to
for uniformity and simplicity of presentation.

>>>>

In our generic version of \AgdaRecord{Semantics}, we keep the
\AgdaField{ren\textasciicircum$\V$} and \AgdaField{$\llbracket$var$\rrbracket$}
fields as before, and replace \AgdaField{$\llbracket$app$\rrbracket$} and
\AgdaField{$\llbracket$lam$\rrbracket$} by a
\AgdaField{$\llbracket$con$\rrbracket$} field as follows.
\[
  \AgdaField{$\llbracket$con$\rrbracket$}~\AgdaBound{r}~\AgdaSymbol{:}~%
  \forallb{%
    \Box\plr{{} \env\V \Delta_1 \dotto {} \sdtstile{}\C A_1} \dottimes
    \cdots \dottimes
    \Box\plr{{} \env\V \Delta_n \dotto {} \sdtstile{}\C A_n} \dotto
    {} \sdtstile{}\C A}
\]
I use ${} \sdtstile{}\C A$ for the Agda notation
\AgdaFunction{\_$\C\vDash$}\AgdaSpace{}\AgdaBound{A}, while ${} \env\V \Delta$
stands for the type family of environments
\AgdaSymbol{$\lambda$}\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}%
\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaFunction{Env}\AgdaSpace{}\AgdaBound{$\V$}%
\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaBound{$\Delta$}.
Environments appear in this definition simply as a way to write a product of
$\V$-values --- one value for each element of $\Delta$.
We could make a special case of premises which do not bind any variables, as did
\citet{AACMM21}, eliding the $\Box$ and empty environment, but I choose not to
for uniformity and simplicity of presentation.

To generate the expressions involving ellipses, I give an interpretation of the
formal rule descriptions.
The interpretation is parametrised on some
\ExecuteMetaData[\SimpleSyntaxtex]{WithScope}, where, in
\AgdaBound{,}\AgdaSpace{}\AgdaBound{$\Delta$}\AgdaSpace{}%
\AgdaBound{$\llbracket$}\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}%
\AgdaBound{$\vdash$}\AgdaSpace{}\AgdaBound{A}\AgdaSpace{}%
\AgdaBound{$\rrbracket$}, the context \AgdaBound{$\Delta$} stands for the newly
bound variables of a premise, \AgdaBound{$\Gamma$} is the context as it was
below the rule's horizontal line, and \AgdaBound{A} is the type of the premise.
In the \AgdaInductiveConstructor{con} constructor for terms, the parameter is
$\Gamma, \Delta \vdash A$, and in the \AgdaField{$\llbracket$con$\rrbracket$}
field for semantics, the parameter is
$\Box\plr{{} \env\V \Delta \dotto {} \sdtstile{}\C A}\,\Gamma$.

A single premise with newly bound variables is interpreted by shuffling the
parts into the right place, while multiple premises are interpreted as pointwise
products of the individual premises (giving the ellipses above).

\ExecuteMetaData[\SimpleSyntaxtex]{semp}

A rule, with all its parameters instantiated, targets a specific type
\AgdaBound{A$'$}, which we check to match the desired type \AgdaBound{A}.
Finally, a whole \AgdaRecord{System} comprises a set \AgdaBound{L} of rule
labels, and \AgdaBound{rs}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}\AgdaBound{L}%
\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaRecord{Rule}.
The interpretation of these data is to pick a rule label \AgdaBound{l}, and then
take the interpretation of the rule \AgdaBound{rs}\AgdaSpace{}\AgdaBound{l}.
For the sake of defining terms as a least fixed point, it is important to note
that the interpretation of a syntax description is strictly positive in the
parameter \AgdaBound{,\_$\llbracket$\_$\vdash$\_$\rrbracket$}.

\ExecuteMetaData[\SimpleSyntaxtex]{semr}
\ExecuteMetaData[\SimpleSyntaxtex]{sems}

The interpretation of a system description as a single layer of syntax is
functorial, supporting the \AgdaFunction{map-s} function when the parameter
\AgdaBound{,\_$\llbracket$\_$\vdash$\_$\rrbracket$} is given as an extra
argument named \AgdaBound{X} or \AgdaBound{Y} (which are both fixed as
parameters of \AgdaFunction{map-s}, together with \AgdaBound{$\Gamma$} and
\AgdaBound{$\Delta$}).

\ExecuteMetaData[\SimpleSyntaxtex]{map-s-type}

The implementation of \AgdaFunction{map-s} is straightforward, so I do not list
it here.
We preserve
the shape of the syntactic layer, applying the function to each \AgdaBound{X}
we find (wherever the description contains \AgdaInductiveConstructor{$\langle$}%
\AgdaSpace{}\AgdaBound{$\Delta$}\AgdaSpace{}%
\AgdaInductiveConstructor{`$\vdash$}\AgdaSpace{}\AgdaBound{A}\AgdaSpace{}%
\AgdaInductiveConstructor{$\rangle$}).

This \AgdaFunction{map-s} will be used in the generic syntax version of
\AgdaFunction{sem} to recursively apply \AgdaFunction{sem} to all subterms.
However, a major distinction between generic syntax and the specific syntax of
previous sections is that the subterms found by \AgdaFunction{map-s} are not
recognised by Agda's termination checker as \emph{structurally smaller} than
the original term.
Therefore, a na\"{i}vely written \AgdaFunction{sem} will fail Agda's termination
check.

To make \AgdaFunction{sem} pass the termination check, we have three major
options:
\begin{enumerate}
  \item Assert \AgdaFunction{sem} to be terminating, bypassing the termination
    check.
  \item Use Agda's \emph{sized types} to remember that the subterms are smaller.
  \item Avoid sized types, and index terms over some user-defined type (for
    example, natural numbers or ordinal notations) which is structurally smaller
    at subterms.
\end{enumerate}

Each of these approaches has drawbacks.
Approach 1 is clearly unsafe, in the sense that the fundamental lemma
\AgdaFunction{sem} is not being completely checked for type-correctness.
Approach 2 is also unsafe, because Agda's sized type implementation is known to
make the system inconsistent~\citep{AgdaIssue1201}.
Meanwhile, approach 3 is safe, but entails a lot of manually extracted and
supplied extra arguments, which I think would distract from the presentation and
make the resulting code harder to use.
I choose to follow \citet{AACMM21} in using sized types, justified by the idea
that Agda may eventually have a sound implementation of sized types, at which
point I would want my code to be as easy to update for that new version of Agda
as possible.

Using sized types, my type family of \AgdaRecord{System}-generic terms is as
below.
\AgdaFunction{Scope} is a name for the transformation of an
\AgdaFunction{OpenFam} into a \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}%
\AgdaSpace{}\AgdaFunction{OpenFam} which appends the extra context to the
existing context before applying the original \AgdaFunction{OpenFam}
(in this case, producing something like $\Gamma, \Delta \vdash A$ from
$\vdash$).
The Agda builtin \AgdaPostulate{$\uparrow$} produces a bigger size from an
existing size \AgdaBound{sz}, giving us here that the size of a term is 1 bigger
than the size of all of its immediate subterms.
Agda's elaborator and termination checker have special support for sizes, so we
do not have to worry much about them from this point on.

\ExecuteMetaData[\SimpleTermtex]{Tm}

Corresponding to the generic \AgdaInductiveConstructor{`con} constructor for
terms, we have a generic field \AgdaField{$\llbracket$con$\rrbracket$} in the
updated \AgdaRecord{Semantics} record.
In place of where one might expect
\AgdaFunction{Scope}\AgdaSpace{}\AgdaBound{$\C$}, we instead have
\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$},
with \AgdaFunction{Kripke} defined below.
The form of \AgdaFunction{Kripke} follows from the shape we saw in the type of
the \AgdaField{$\llbracket$lam$\rrbracket$} field we saw in \cref{sec:gen-sem},
where I use an environment targeting \AgdaBound{$\Delta$} as a way to say
``a value for each type in \AgdaBound{$\Delta$}''.

\ExecuteMetaData[\SimpleSemanticstex]{Kripke}
\ExecuteMetaData[\SimpleSemanticstex]{Semantics}

Finally, we get a generic semantic traversal as follows.
The function \AgdaFunction{bindEnv} is unchanged from \cref{sec:gen-sem}, as it
never mentions the syntax.
The type of the traversal \AgdaFunction{sem} is also basically unchanged --- we
just need to account for arbitrary term sizes (\AgdaBound{sz}), which will get
smaller when recursing on subterms.
I have chosen, as did \citet{AACMM21}, to define \AgdaFunction{sem} mutually
with a function \AgdaFunction{body}, which is like a counterpart to
\AgdaFunction{sem} dealing with newly bound variables.
Note that the mutual recursion is not essential --- for example,
\AgdaFunction{body} could simply be inlined.
The \AgdaInductiveConstructor{`var} case of \AgdaFunction{sem} is as before.
The \AgdaInductiveConstructor{`con} case, if viewed appropriately, is a direct
generalisation of the \AgdaInductiveConstructor{lam} case from earlier.
We recursively apply \AgdaFunction{sem} to all immediate subterms contained in
\AgdaBound{M} (as found by \AgdaFunction{map-s}), with an environment updated to
reflect the newly bound variables in each premise of the rule that was applied.

\ExecuteMetaData[\SimpleSemanticstex]{sem}

### 2023-02-20 tex/mech-related.tex
[ writing ] new draft of mechanisation chapter

\subsection{Fiore, Plotkin, Turi}

TODO: cite \citet{FPT99} and \citet{FS22}.

>>>>

\subsection{Fiore et al.}

\todo{This section is currently a random selection of paragraphs.}
Marcelo Fiore and various collaborators have a long line of work aiming for a
categorical account of variable-binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
A recent, particularly relevant paper from this line of work is \citet{FS22},
which mechanises some of this work to obtain a framework similar in scope to
the work of \citet{AACMM21}.

Perhaps the main methodological difference between the works by
\citet{AACMM21} and \citet{FS22} is that \citet{AACMM21} start from typed
syntaxes and then give them semantics, whereas \citet{FS22} start by
characterising the category of models of a theory, and then later showing that
syntaxes form the initial model.
The latter approach means that \citeauthor{FS22} can do a significant amount of
theory not mentioning syntax in a compositional and structured style.
In contrast, the former approach is more lightweight when one wants to write
programs without a categorical specification.

I chose to base the work of this thesis on that of \citet{AACMM21} for largely
circumstantial reasons.
I only became aware of the work of \citet{FS22} when it was published, by which
time I had already completed the bulk of the work presented in this thesis.
Also, one of the authors of the former paper is my PhD supervisor, and the other
authors too were working nearby, so it was easy to discuss their work quickly
and informally.

In terms of the underlying theory, both works extend multi-sorted universal
algebra\todo{cite} with variable-binding.
However, the two extensions are subtly different.
Universal algebra already has a notion of variable, which supports renaming and
substitution, and which allows a given term to be evaluated when the free
variables of that term are assigned semantic values.
\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
variables in terms.
On the other hand, \citeauthor{FS22} recast the existing variables as
\emph{metavariables}, and introduce a new notion of (bindable) variables
separately into the syntax.
The resulting metavariables can then stand for arbitrary \emph{open} terms, and
thus each one remembers its context and requires a substitution whenever it is
used.

%Compared to the work of \citet{AACMM21}, which is what I chose to guide this
%chapter and the rest of this thesis, \citet{FS22} can be understood as providing
%many extra features.
%I happened not to be interested in these specific features, so have not
%implemented them, but future development of the work I present in this thesis
%could benefit from such features.
%The extra features include an externally compiled language of syntax and theory
%descriptions, internal support for metavariables, and a treatment of the
%category of models (of which terms are the initial model and what I call
%\AgdaFunction{sem} forms the universal morphism from terms to any other model).

\Citeauthor{FS22} use metavariables to form descriptions of relations over
terms, like an equational theory over the simply typed $\lambda$-calculus.
Terms have, as well as their context of variables, a context of metavariables,
and terms can contain a metavariable wherever they could contain a subterm.
There is then an operation of metavariable substitution, which substitutes terms
in the place of metavariables.
Metavariable substitution is used to instantiate the rules of described
relations/theories.
In contrast, \citeauthor{AACMM21} make do with the variables of the metalanguage
(i.e.\ Agda variables).
\todo{Congruence once and for all.}

Technologically, \citeauthor{FS22} provide an external domain-specific language
for syntax descriptions.
From such a description, a Python program generates some boilerplate Agda code
providing the types, the algebraic signature, the well typed terms, and a proof
that the terms are the initial model.
Using code generation like this resembles parts of the Autosubst plugin, as
opposed to the purely internal, generic programming-based solution given by
\citeauthor{AACMM21} and in this thesis.

\Citet[p.\ 19]{FS22} avoid sized types by defining their equivalent of the
functorial mapping \AgdaFunction{map-s} specialised to \AgdaFunction{sem}, with
these two functions being mutually recursive.
This is a standard solution, where code reuse is traded away in favour of
satisfying the termination checker and avoiding sized types.

### 2023-02-20 tex/terms.tex
[ edit ] some corrections and notes after meeting
[ writing ] new draft of mechanisation chapter

Most mechanisations choose contexts to be an inductive list of types.
However, I instead choose a functional, tree-shaped representation, as shown
with the type \AgdaRecord{Ctx}.
The type \AgdaDatatype{LTree} is the inductive type generated by leaves and
nullary \& binary nodes, and serves as a generalised ``length'' of the context.
The contents of the context --- the types --- are then stored in the functional
vector \AgdaField{ty-ctx}, which is a mapping from leaves in \AgdaField{shape}
to object language types \AgdaDatatype{Ty}.
The advantages of the functional vector representation will not become clear
until later chapters\todo{forward reference}, where I make use of the ease of
look-up and $\eta$-law of functions.
However, I claim for now that there is little to no disadvantage in the
functional vector representation --- in particular, we have no need for
function extensionality principles because we never talk about equality of
contexts.
For example, instead of using an equality of contexts to coerce a term, we can
use renaming.
As for the tree shape, this makes context concatenation definitionally
injective, so that in cases where multiple new variables are bound in a subterm
(for example, $\otimes$-elimination), Agda's unification-based solving will
be more able to infer which variables have just been bound.

@@

>>>>

Most mechanisations choose contexts to be an inductive list of types.
However, I instead choose a functional, tree-shaped representation, as shown
with the type \AgdaRecord{Ctx}.
The type \AgdaDatatype{LTree} is the inductive type generated by leaves and
nullary \& binary nodes, and serves as a generalised ``length'' of the context.
The contents of the context --- the types --- are then stored in the functional
vector \AgdaField{ty-ctx}, which is a mapping from leaves in \AgdaField{shape}
to object language types \AgdaDatatype{Ty}.
The advantages of the functional vector representation will not become clear
until later chapters --- particularly the example in
\cref{sec:usage-elaborator}, where I make use of the ease of look-up and
$\eta$-law of functions.
However, I claim for now that there is little to no disadvantage in the
functional vector representation --- in particular, we have no need for
function extensionality principles because we never talk about equality of
contexts.
For example, instead of using an equality of contexts to coerce a term, we can
use renaming.
As for the tree shape, this makes context concatenation definitionally
injective, so that in cases where multiple new variables are bound in a subterm
(for example, $\otimes$-elimination), Agda's unification-based solving will
be more able to infer which variables have just been bound.

@@

\todo{Example terms}
